Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-04-01 paper/manuscript.tex
first draft of atom mapping update

% aux. comments Stefan Boresch -> Stefan Boresch
\newcommand{\sbnote}[1]{%
  {\bfseries{}[SB: }%
  {\textcolor{blue}{#1}}{\bfseries{}]}
  }
\newcommand{\rref}{{\bfseries[REFs]}{}}

@@

\paragraph{Topologies} In regular biomolecular simulations one studies physical systems; in particular, all entities in the system, e.g., protein, ligand and waters, represent real molecules. For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology. In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc. (see {\bf XXX}). The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages. In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B. The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state. We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates. The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously. Thus, we have dual coordinates, but chemical topologies for each of the entities. In the following, we outline these approaches in more detail, together with additional considerations how the various approaches in practical use differ. While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation hardware.

\paragraph{Single coordinate approaches --- alchemical topologies}

When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules. %Often this is predetermined by the choice of simulation software.
As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}. The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.

@@

\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.

\paragraph{Dual coordinate approaches --- dual chemical topologies}

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{Ssee below; where do we place the dummy atom discussion?}. Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

Dual coordinate approaches with chemical topologies encompass a third approach, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

%Other methodologies also use both end states simultaneously; examples are the Separated Topologies (SepTop) approach [Baumann2023] and the ATM method [Azimi2022]. In the former, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to. Ries et al. refer to such an approach as “separated dual topology”, whereas they label their approach as “linked dual topology”. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space.

%One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.

\paragraph{Summary, comparison of single and dual coordinate methods, orthogonal classifications:}

%Some additional terms have also been employed to talk about these different intermediate pathways . Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.

Unfortunately, during the past years the use of the terms single and dual topology has proliferated, often with quite different meaning from the original intentions. The above presentation makes clear that the primary difference between approaches is the use of separate coordinates for each end state (chemical topology) or the use of an alchemical chimera with a single set of coordinates (alchemical topology). We therefore suggest the following classification
\begin{itemize}
\item Single coordinate (SC) --- Alchemical topology
  \begin{itemize}
  \item Single topology (SCST)
    \item Dual topology (SCDT)
  \end{itemize}
  \item Dual coordinate --- Chemical topology (DC)
  \end{itemize}
For completeness, we list some additional terminology that has been used: Some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though in our view these calculations should be classified as single coordinate single topology \sbnote{Check whether this is correct!}. %this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
Ries et al.\ suggested terminology \cite{Ries_2022} in which single topology corresponds to SCST, hybrid topology to SCDT, and dual topology to DC.

The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology. In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \rref or restraints. In the latter case, the restraints may couple the two set of coordinates directly \rref, or the restraints may be defined relative to a common reference frame. Finally, in ATM the two copies of the coordinates are held separate.

Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~{\bfseries XX}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters. E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$. For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for an aliphatic C-H bond, and for $\lambda=1$ to those for an aliphatic C-C bond. The other energy terms are treated analogously. The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.  More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.

The $\lambda$-dependent interpolation can also be accomplished by mixing the energies and forces of the initial ($\lambda=0$) and final state ($\lambda=1$), e.g., linearly according to $U(\lambda)=(1-\lambda) U_0+\lambda\,U_1$. Other λ-dependencies can be used, and soft-cores \rref may be used for Lennard-Jones and electrostatic interactions to avoid endpoint problems. Giese et al. \cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}.

Any SCDT approach requires two energy/force calculations and, thus, uses energy mixing. Therefore, approaches that implement energy mixing together with an alchemical topology (a single set of coordinates) often support both SCST and SCDT. E.g., CHARMM's PERT module just mentioned could be used to set up either of the SC approaches sketched in Fig.~{\bf XX}. Finally, all DC approaches require two energy/force evaluations.

Any energy mixing approach, whether in a SC or DC context, requires two energy/force evaluations. While various optimizations keep the computational overhead significantly below a factor of two,  parameter mixing, if applicable, is the compuationally most efficient approach. However, it is closely tied to SCST and the more complicated the alchemical transformation, the more challenging a strict SCST approach becomes. Also, the mixing of partial charges and the calculation of $\partial U/\partial\lambda$ for the reciprocal space particle-mesh Ewald (PME) summation is tricky. Giese et al.\ described how to do this \cite{Giese_2018}, but most energy mixing implementations calculate the PME term twice and mix the resulting energy and forces.

\sbnote{I think this should get its own paragraph/subsection; to be expanded!} Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach. Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both). 
To our knowledge, efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.

\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed, meaning there is no MCSS match between the two compounds.}
    \label{fig:fig_mcss}
\end{figure}

There are different tools that allow the generation of MCSS matches as well as single topology input. A large number of software tools can compute MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}. The MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space. Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode. If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.

Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem. Using an absolute free energy approach instead seems more promising in such a case.

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.

@@

\section*{Author Information}
\makeorcid
\bibliography{alchemical,manual,sb}

>>>>

@@

\paragraph{Topologies} 
A critical first step is to determine which approach to use for atom mapping between end state molecules. Often this is predetermined by the choice of simulation software. Typically it is possible to chose between a \emph{dual topology} versus a \emph{single topology}

The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.

@@

As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.

Some additional terms have also been employed to talk about these different intermediate pathways . Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.

One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.

Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach. Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both). 
To our knowledge, efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.

\paragraph{Atom mapping}
Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping. This is the case especially for system representations using a single coordinate set (Type 1. & 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbedbetween the two molecules (respectivly end-states).
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done. This however requires either careful implementation of atom mass changes, bond length etc. or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors. \cite{ries2024Kartograf}

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed, meaning there is no MCSS match between the two compounds.}
    \label{fig:fig_mcss}
\end{figure}

In general there are mutliple ways on how a MCSS can be calculated. One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
A large number of software tools can compute SIS MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry. Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}. The SIS MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.

Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules. These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g. stereochemistry or binding mode differing).
The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned. Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or kartograf \cite{ries2024kartograf}.

In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode. If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.

A next step for atom mappings, would be to support multistate methods (e.g. RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms. There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect. \cite{petrov2021Perturbation}

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation. Few descriptions of the approach can be found for such approaches like the one suggested by Baumann et al.~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute} for seperated topology approach, finding , or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, Rieder2022Lev}
Alternativley an approach where such a restraint algorithm was included in a pipeline for ABFE calculations is described by the ABFE\_workflow \cite{Ries2024ABFE} using MDRestraintsGenerator\cite{alibay2021mdrestraintsgenerator}.

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address thisusing a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.

@@

\section*{Author Information}
\makeorcid
\bibliography{alchemical,manual}
